How Kids Code and How We Know:\\An Exploratory Study on the Scratch Repository

### 2016-06-15 Paper/paper.tex
Intergrated Felienne's comments

Block-based languages have existed since the eighties \cite{e._glinert_towards_1986}, but have recently found adoption as tools for programming education. In addition to Scratch, also Alice~\cite{conway_alice:_1994}, Blockly\footnote{\url{https://developers.google.com/blockly/}} and App Inventor~\cite{wolber_app_2011} are block-languages aimed at novice programmers.

Several studies have shown that block-based languages are powerful as tools for teaching programming~\cite{meerbaum-salant_learning_2010, b._moskal_evaluating_2005,cooper_teaching_2003,price_comparing_2015}. Previous works involving static analysis of Scratch programs have evaluated the application of various programming concepts in Scratch projects \cite{Maloney_2008,moreno_automatic_2014}. Recent works have focused on bad programming practices within Scratch programs \cite{Meerbaum_habits_2011}, and automated quality assessment tools have been proposed for identifying code smells~\cite{fowler_refactoring:_1999}~and bad programming practices \cite{boe_hairball:_2013,moreno_automatic_2014}. In a recent controlled experiment we found that long scripts and code duplication decreases a novice programmer's ability to understand and modify Scratch programs \cite{hermans_2016}.

The goal of this paper is to obtain an understanding of how people program in Scratch by analyzing their programming artifacts, i.e., the saved and shared Scratch projects, and to quantitatively evaluate the use of programming abstractions and concepts. Moreover, knowing that bad programming habits and code smells can be harmful \cite{hermans_2016}, we also want to explore whether they are common. To address this goal, we answer the following research questions:

@@

Our study is based on data from the Scratch project repository. By scraping the list of recent projects,\footnote{\label{scratchpublic}\url{https://scratch.mit.edu/explore/projects/all/}} we have obtained \nPrograms~public Scratch projects and performed source code analysis on them. To the best of our knowledge, this is the first large-scale exploratory study of Scratch programs.

@@

\section{Relevant Scratch Concepts}
\label{sec:scratch}
This paper is by no means an introduction into Scratch programming, we refer the reader to \cite{brennan_creative_2014} for an extensive overview. To make this paper self-contained, however, we explain a number of relevant concepts in this section.

@@

\begin{description}
	\item [Sprites] Scratch code is organized by `sprites': two dimensional pictures each having their own associated code. Scratch allows users to bring their sprites to life in various ways, for example by moving them in the plane, having them say or think words or sentences via text balloons, but also by having them make sounds, grow, shrink and switch costumes. The Scratch program in Figure \ref{fig:ui} contains two sprites, the cat, which is Scratch's default sprite and a piano. The code in Sprite1 will cause the cat to move right when the right arrow is pressed, and when the green flag is clicked it will continuously sense touching the piano.
	\item [Scripts]Sprites can have multiple code blocks, called scripts. The Scratch code in Figure \ref{fig:ui} has two distinct scripts, one started by clicking on the green flag and one by pressing the space bar. It is possible for a single sprite to have multiple scripts initiated by the same event. In that case, all scripts will be executed simultaneously.
	\item [Events] Scratch is \emph{event-driven}: all motions, sounds and changes in the looks of sprites are initiated by events called Hat blocks\footnote{\url{http://wiki.scratch.mit.edu/wiki/Hat_Block}}). The canonical event is the \texttt{when Green Flag clicked}, activated by clicking the green flag at the top of the user interface. In addition to the green flag, there are a number of other events possible, including key presses, mouse clicks and input from a computer's microphone or webcam. The Scratch code in Sprite1 in Figure \ref{fig:ui} contains two events: \texttt{when Green Flag clicked} and \texttt{when <right arrow> key pressed}, each with associated blocks.
	\item [Signals] Events within Scratch can be user generated too: users can broadcast a message, for example when two sprites touch each other, like in Figure \ref{fig:ui}. All other sprites can then react by using the \texttt{when I receive} Hat block. In Figure \ref{fig:ui}, Sprite1 broadcasts `bump' when the cat touches the Piano. 
	\item[Custom blocks] Scratch users can define their own blocks, which users can name themselves, called custom blocks. The creation of custom blocks in Scratch is the equivalent of defining procedures in other languages \cite{moreno_automatic_2014}. Because the term `procedures' is common in related work, we will refer to custom blocks as `procedures' in the remainder of this paper. Procedures can have input parameters of type string, number, and boolean. When a user defines a procedure, a new Hat block called \texttt{define} appears, which users can fill with the implementation of their block.
\end{description}

@@

The main focus of this study is to understand how people program in Scratch by analyzing the characteristics of Scratch projects. To answer our three research questions, we conducted an empirical quantitative evaluation of project data we collected from the Scratch project repository. In the following paragraphs we describe the dataset, the process and the tools we used for analyzing it, and the methods we followed for detecting code smells.

@@

We obtained the set of Scratch projects by using a scraping program. Our scraping program called Kragle starts by reading the Scratch projects page\footnotemark[\ref{scratchpublic}] and thus obtains project ids of projects that were most recently shared. Subsequently, Kragle retrieves the JSON code for each of the listed projects.

@@

We ran Kragle on March 2nd 2016 for 24 hours and, in that time, it obtained a little over 250,000 projects. Out of the \nPrograms, we failed to parse and further analyze 2,367 projects due to technical difficulties with the provided JSON files. Kragle, as well as all scraped projects and our analysis files are available.\footnote{\label{repo}\url{https://github.com/TUDelftScratchLab/ScratchDataset}}

Once we obtained the Scratch projects, we parsed the JSON files according to the specification of the format.\footnote{\url{http://wiki.scratch.mit.edu/wiki/Scratch_File_Format_(2.0)}} This resulted in a list of used blocks per project, with the sprites and the stage of the project. We cross referenced also all blocks with the Scratch wiki to determine the shapes and the category of all blocks. For example, \texttt{When Green Flag Clicked} is a \emph{Hat block} from the \emph{Events category}.

@@

For RQ1 we measured the size of projects based on the number of blocks in scripts and sprites and we calculated descriptive statistics, which are presented in Section \ref{RQ1}. For measuring the complexity of the scripts we used the McCabe cyclomatic complexity metric \cite{mcCabe76}, a quantitative measure of the number of independent paths through a program's source code. This is calculated per script by counting the number of decision points in the script plus one. In Scratch, decision points can be the \texttt{if} and \texttt{if else} blocks.

For RQ2, we used the data on the code blocks and their categories to perform statistical analysis of applied programming abstractions and concepts. Similarly to \cite{Maloney_2008}, we consider the use of certain blocks to indicate that a programming abstraction or concept is being used in a certain project. In Section \ref{RQ2} we present the results related to the utilization of procedures, variables, loops, conditional statements, user interactivity and synchronization.

For RQ3, we focused on four types of code smells: duplicated code, dead code, large script and large sprite. For the duplicated code smell analysis, our first step was to specify what we consider a code clone in the context of Scratch programming: a script that is composed of a set of blocks of the same type connected in the same way and is repeated within or across sprites of the same project. For the identification of clones we did not take into account the values of the parameters that may be used in the blocks, so that two blocks that only differ in the values of parameters are considered to be equal. We also examined the case of clones with the same parameter values, and we refer to them as \textit{exact clones}. The next step in the analysis was to determine the minimum size of the scripts that are considered clones instead of incidentally similar. We examined the number of detected clones for different script sizes and present the results in Figure \ref{fig:cloneslines}. Based on this distribution, we opted to adopt the number also used by the authors in \cite{moreno_automatic_2014}, which is the minimum size of 5 blocks per script.

@@

To examine the long method and the large class smells, we consider them in the context of Scratch as large script and large sprite smells respectively. For these two smells we use the number of blocks as the size metric. Figure \ref{fig:longmethod} presents the number of blocks in the scripts and the sprites of our dataset. We used these numbers to split the dataset and retrieve the top 10\% largest scripts and sprites, as is commonly done in both source code analysis~\cite{alves_deriving_2010} and analysis of end-user programming artifacts like spreadsheets~\cite{hermans2015detecting}. Using that strategy, we set the thresholds for the calculation of the large script and large sprite: it is 18 blocks and 59 blocks respectively. The results we obtained using these thresholds are presented in Section \ref{RQ3}.

@@

\subsection{Program Size and Complexity}
\label{RQ1}
The dataset contains a relatively small number of projects without any code: \nemptyPrograms~ (5.77\%). Through random manual sampling we found that in some cases these projects contains only sprites and costumes, but no code, while in other projects they were entirety empty apart from the Scratch cat added by default. Since these projects are empty in terms of code we excluded them from further analysis, leaving the final number of analyzed non-empty projects to \nScriptPrograms.

\begin{table*}[ht]
	\centering
	\begin{tabular}{llrrrrrr}
		&&\textbf{mean}&\textbf{min}&\textbf{Q1}&\textbf{median}&\textbf{Q3}&\textbf{max}\\
		\hline
		Size&Sprites with code per project&5.68&1&1&2&5&525\\
		&Scripts per project&17.35&1&2&5&12&3,038\\
		&Number of blocks per project&154.55&1&12&29&76&34,622\\
		&Blocks in Stage per project&4.80&0&0&0&3&2,613\\
		&Blocks in Sprites per project&115.57&0&10&26&68&34,613\\
		&Blocks in Procedures per project&34.17&0&0&0&0&20,552\\
		\hline
		Complexity&McCabe Cyclomatic Complexity (CC) per script&1.58&1&1&1&1&246\\
		&McCabe CC per procedure script&3.75&1&1&2&4&183\\
		\hline
		Procedures&Procedures per project with procedures&11.50&1&1&2&6&847\\
		&Arguments per Procedure&0.95&0&0&0&1&53\\
		&Numerical arguments per procedure with arguments&1.73&0&1&1&2&22\\
		&Text arguments per procedure with arguments&0.28&0&0&0&1&24\\
		&Boolean arguments per procedure with arguments&0.13&0&0&0&0&14\\
		&Calls per procedure&2.14&0&1&1&2&526\\
		&Scripts with calls per procedure&1.13&0&1&1&1&59\\
		\hline
		Programming&Variables per project&2.06&0&0&0&1&340\\
		concepts&Scripts utilizing variable&4.97&1&1&3&5&1,127\\
		&Lists per project&0.55&0&0&0&0&319\\
		&Conditional statements per project&10.02&0&0&0&3&5,950\\
		&Loop statements per project&7.65&0&1&2&5&2,503\\
		&User input blocks per project&4.77&0&0&1&4&1,889\\
		&Broadcast-receive statements per project&8.57&0&0&0&2&2,460\\	
		\hline
	\end{tabular}
	\caption{Summary statistics from the dataset of \nScriptPrograms~non-empty Scratch projects}
	\label{tbl-size}
\end{table*}

@@

The number of blocks metric was further analyzed to understand code organization. The majority of Scratch code---74.78\% out of \nLOC~blocks---is written within sprites. An additional 3.1\% of the total blocks are found in the stage class. More interestingly, the remaining 22.11\% are blocks within defined procedures, which are found in only 7.7\% (17,979) of the projects. The projects that contain procedures use them a lot; almost half of their total blocks (48.81\%) are within procedures.

@@

\noindent
\fbox{
	\begin{minipage}{0.46\textwidth}
		RQ1: The majority of Scratch projects are small and simple; 75\% of the projects have up to 5 sprites, 12 scripts and 76 blocks. The majority (78\%) of scripts contain no decision points. Most code is written in sprites. There exist surprisingly large and complex projects.
	\end{minipage}}

\subsection{Programming Abstractions and Concepts}
\label{RQ2}
The first method for abstraction that we investigate are procedures. In the dataset we found 206,799 procedures in 17,979 (7.7\%) projects. As summarized in Table \ref{tbl-size}, the projects that contain procedures have an average of 11.5 procedures, but with 53.59\% of these projects having up to 2. Figure \ref{fig:proceduresperproject} shows the distribution of procedures in projects. Regarding procedure arguments, we found that 55.57\% have no arguments and 19.48\% have only one (shown in Figure \ref{fig:procedurearguments}). The majority of procedure arguments (80.59\%) are numeric, and the least used argument type is the boolean one---6.23\% of the total procedure arguments, found in 5.32\% of the procedures.

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/6proceduresperproject}
	\vspace{-1em}
	\caption{Number of procedures for the 17,979 projects that include at least one}
	\label{fig:proceduresperproject}
\end{figure}

@@

The use of procedures in projects was further investigated through the procedure calls, summarized in Figure \ref{fig:procedurecalls}. Most procedures are called exactly once (62.32\% of them) or twice (14.30\%) and from exactly one script (85.92\% of them). Examining the origin of procedure calls, we observed that most of the calls (56.09\%) originate from other procedures, and even 1.06\% originate from the same procedure, making them recursive calls. These recursive procedures are found in 1,052 projects, whose identifiers are made available.\footnotemark[\ref{repo}]

@@

As shown in Table \ref{tbl-characteristics}, almost one-third of the projects use variables and a small number (4.01\%) use lists. The number of variables that is being used is also limited, with only 7.48\% of the projects having 5 or more variables. The distribution of variable and list utilization is shown in Figure \ref{fig:variables}. Exceptional cases exist: the analysis highlighted 842 projects with more than 100 variables and with a maximum of 340. Examining the initialization of variables through the \texttt{set <variable> to <value>} blocks, we found that for 4.83\% of all variables this was missing. While failing to initialize a variable in Scratch will not result in a runtime error as in some other programming languages, correctly setting the initial state of the program is important \cite{boe_hairball:_2013}.

@@

Regarding program control features, conditional statements (blocks \texttt{if <condition> then} and \texttt{if <condition> then else}) are used by 39.81\% of the projects. Loops (blocks \texttt{repeat <times>}, \texttt{forever} and \texttt{repeat until <condition>}) are more common, used by 77.18\% of the projects. The most common of the three is the \texttt{forever} block, accounting for 51.86\% of all loops and the least common one is the \texttt{repeat until <condition>} block, accounting for 11.57\% and used in 13.59\% of the total projects.

Investigating user interactivity functionality, we found that 56.24\% of the projects in the dataset contain user input blocks---an average of 8.48 blocks per such project. Table \ref{tbl-userinput} lists the frequency of use of user input controls. We do not include the \texttt{when Green flag pressed} block here, as this is just used to start a Scratch program and hence cannot really be considered input into the program. The most commonly used user input block is the \texttt{when key pressed}, found in 71,096 (30.45\% of the total) projects. The most frequently used parameter for the \texttt{key} attribute is the space key, followed by the arrows and then the letters and numbers.

@@

Users can define their own events, using the blocks \texttt{broadcast}, \texttt{broadcast and wait} and \texttt{when I receive}.These blocks are used by 29.57\% of the projects. \texttt{broadcast and wait} is rarely used, in only 3.87\% of the projects.

@@

We further inspected which of the identified clones were duplicated only within the same sprite: 63,682 (37.34\% of the total) clones, in 10.14\% of all projects. Procedure clones were measured to 12,878 (7.55\% of the total) clones, in 2.12\% of the projects.

@@

Apart from whole scripts we also examined cases where scripts differed only in the first (Hat) block. This way we examine if Scratch programmers assign the same functionality to handle different types of events. Cloned functionality blocks are found to be rare: without considering the first block, only 2,243 additional clones were found in 920 projects.

The second smell that we examine is the dead code smell. We identify four types of dead code: (1) procedures that are not invoked, (2) unmatched broadcast-receive messages, (3) code that is not invoked and (4) empty event scripts. Investigating the first type, we find that a significant number of the defined procedures (13,036 or 5.06\%) are not called in the projects. This is also shown in Figure \ref{fig:procedurecalls} and it occurs in 2,079 projects. For the second type, we examined the broadcast-receive messages and found that they are not always synchronized: 3,33\% of the \texttt{when I receive} blocks were found to wait for a message that is never being broadcasted, while 4,4\% of the \texttt{broadcast} blocks broadcast a message that is not being received. This lack of syncronization occurs in 18,669 (7.99\% of the total) projects.

The third and fourth cases are incomplete scripts. They are either never invoked due to the lack of a starting \texttt{when <trigger>} block from the Scratch Events or Control category, or are comprised of only a \texttt{when <trigger>} block without any functionality. A total of 322,475 scripts like that were found in 56,890 (24.36\% of the total) projects. The majority of these scripts (86.6\%) are scripts missing the starting block. Examining the size of these dead scripts, 72.34\% are composed of a single block. As shown in Figure \ref{fig:deadcode}, however, considerably large dead scripts exist; 2,358 of these scripts originating from 1,553 different projects have more than 30 blocks and up to 2,610.\footnotemark[\ref{repo}]

@@

\section{Discussion}
\label{sec:discussion}
\subsection{Implications}
We believe a large scale study of programs like ours can help language designers to tailor their language. In this section we highlight directions in which our study could support language design. There are many other implications to be considered, which is why we have made our dataset public.

@@

In a sense, the dead scripts are harmless, as they are not executed. However, they do cause `visual clutter' and might be distracting to novice programmers, as it might be hard to see which scripts are dead. In contrast to other visual educational languages, Scratch does not indicate scripts that are dead. LEGO Mindstorms, for example, does give the user such feedback by making unconnected blocks gray.

Looking at the number of unconnected blocks, we hypothesize that Scratch programmers have a need for a separate workspace to store unconnected blocks temporarily. We envision that would be like the `backpack' meant to move scripts across sprites. In order to help novice programmers keep their code clean, the programming interface could actively encourage users to move unconnected blocks to that workspace when they exit the environment.

\subsubsection{Exact clones between sprites}
\label{implicationsClones}
With occurrences in 11\% of the Scratch projects in our dataset, the use of exactly identical clones between sprites is relatively common. In a sense, the Scratch users are not to blame here, as Scratch does not support procedure calls between sprites, only within them. So in many cases there is no way to share the functionality other than by making a copy. We are not aware of the underlying rationale of the Scratch team that lead to this decision, however it seems that a large part of the Scratch users would use the functionality to call procedures between sprites.

@@

\subsection{Threats to validity}
A threat to the validity of this study is the fact that we did not scrape a random sample, but the most recent 250,000 projects. It could be the case that the programming habits of Scratch users are changing over time. However, we counterbalanced that by using a large dataset which comprises of around 2\% of all 14 million shared Scratch projects.\footnote{\url{https://scratch.mit.edu/statistics/}}.

>>>>

Block-based languages have existed since the eighties \cite{e._glinert_towards_1986}, but have recently found adoption as tools for programming education. In addition to Scratch, also Alice~\cite{conway_alice:_1994}, Blockly\footnote{\url{https://developers.google.com/blockly/}} and App Inventor~\cite{wolber_app_2011} are block-based and aimed at novice programmers.

Several studies have shown that block-based languages are powerful as tools for teaching programming~\cite{meerbaum-salant_learning_2010, b._moskal_evaluating_2005,cooper_teaching_2003,price_comparing_2015}. Previous works involving static analysis of Scratch programs have evaluated the application of various programming concepts in Scratch projects \cite{Maloney_2008,moreno_automatic_2014}. Recent works have focused on bad programming practices within Scratch programs \cite{Meerbaum_habits_2011}, and automated quality assessment tools have been proposed for identifying code smells~\cite{fowler_refactoring:_1999}~and bad programming practices \cite{boe_hairball:_2013,moreno_automatic_2014}. In a recent controlled experiment we found that long scripts and code duplication decrease a novice programmer's ability to understand and modify Scratch programs \cite{hermans_2016}.

The goal of this paper is to obtain an understanding of how people program in Scratch by analyzing their programming artifacts, i.e., the saved and shared Scratch projects. With this we aim to quantitatively evaluate the use of programming abstractions and concepts. Moreover, knowing that bad programming habits and code smells can be harmful \cite{hermans_2016}, we also want to explore whether they are common. To address this goal, we answer the following research questions:

@@

Our study is based on data from the Scratch project repository. By scraping the list of recent projects\footnote{\label{scratchpublic}\url{https://scratch.mit.edu/explore/projects/all/}}, we have obtained \nPrograms~public Scratch projects and performed source code analysis on them. To the best of our knowledge, this is the first large-scale exploratory study of Scratch programs.

@@

\section{Relevant Scratch Concepts}
\label{sec:scratch}
This paper is by no means an introduction into Scratch programming. We refer the reader to \cite{brennan_creative_2014} for an extensive overview. To make this paper self-contained, however, we explain a number of relevant concepts in this section.

@@

\begin{description}
	\item [Sprites] Scratch code is organized by `sprites': two dimensional pictures each having their own associated code. Scratch allows users to bring their sprites to life in various ways, for example by moving them in the plane, having them say or think words or sentences via text balloons, but also by having them make sounds, grow, shrink and switch costumes. The Scratch program in Figure \ref{fig:ui} contains two sprites, the cat, which is Scratch's default sprite and a piano. The code in Sprite1 will cause the cat to move right when the right arrow is pressed, and when the green flag is clicked it will continuously sense touching the piano.
	\item [Scripts]Sprites can have multiple code blocks, called scripts. The Scratch code in Figure \ref{fig:ui} has two distinct scripts, one started by clicking on the green flag and one by pressing the space bar. It is possible for a single sprite to have multiple scripts initiated by the same event. In that case, all scripts will be executed simultaneously.
	\item [Events] Scratch is \emph{event-driven}: all motions, sounds and changes in the looks of sprites are initiated by events called Hat blocks\footnote{\url{http://wiki.scratch.mit.edu/wiki/Hat_Block}}). The canonical event is the \texttt{when Green Flag clicked}, activated by clicking the green flag at the top of the user interface. In addition to the green flag, there are a number of other events possible, including key presses, mouse clicks and input from a computer's microphone or webcam. The Scratch code in Sprite1 in Figure \ref{fig:ui} contains two events: \texttt{when Green Flag clicked} and \texttt{when <right arrow> key pressed}, each with associated blocks.
	\item [Signals] Events within Scratch can be user-generated too: users can broadcast a message, for example when two sprites touch each other, like in Figure \ref{fig:ui}. All other sprites can then react by using the \texttt{when I receive} Hat block. In Figure \ref{fig:ui}, Sprite1 broadcasts `bump' when the cat touches the Piano. 
	\item[Custom blocks] Scratch users can define their own blocks, which users can name themselves, called \textit{custom blocks}. The creation of custom blocks in Scratch is the equivalent of defining procedures in other languages \cite{moreno_automatic_2014}. Because the term `procedures' is common in related work, we will refer to custom blocks as `procedures' in the remainder of this paper. Procedure parameters can be textual, numeric or boolean. When a user defines a procedure, a new Hat block called \texttt{define} appears, which users can fill with the implementation of their procedure.
\end{description}

@@

The main focus of this study is to understand how people program in Scratch, by analyzing the characteristics of Scratch projects. To answer our three research questions, we conducted an empirical quantitative evaluation of project data we collected from the Scratch project repository. In the following paragraphs we describe the dataset, the process and the tools we used for analyzing it, and the methods we followed for detecting code smells.

@@

In order to collect the set of Scratch projects in the dataset we built a web scraping program. The scraping program, called Kragle, starts by reading the Scratch projects page\footnotemark[\ref{scratchpublic}] and thus obtains the project identifiers of projects that were most recently shared. Subsequently, Kragle retrieves the JSON code for each of the listed projects.

@@

We ran Kragle on March 2nd 2016 for 24 hours and, during that time, it obtained a little over 250,000 projects. Out of the \nPrograms, we failed to parse and further analyze 2,367 projects due to technical difficulties with the JSON files. Kragle, as well as all scraped projects and our analysis files are available.\footnote{\label{repo}\url{https://github.com/TUDelftScratchLab/ScratchDataset}}

Once we obtained the Scratch projects, we parsed the JSON files according to the specification of the format\footnote{\url{http://wiki.scratch.mit.edu/wiki/Scratch_File_Format_(2.0)}}. This resulted in a list of used blocks per project, with the sprites and the stage of the project. We also cross referenced all blocks with the Scratch wiki to determine the shapes and the category of all blocks. For example, \texttt{When Green Flag Clicked} is a \emph{Hat block} from the \emph{Events category}.

@@

To answer RQ1, we measured the size of projects based on the number of blocks in scripts and sprites and we calculated descriptive statistics, which are presented in Section \ref{RQ1}. For measuring the complexity of the scripts we used the McCabe cyclomatic complexity metric \cite{mcCabe76}, a quantitative measure of the number of independent paths through a program's source code. This is commonly calculated per script by counting the number of decision points in the script plus one. The decision points that Scratch supports are the \texttt{if} and the \texttt{if else} blocks.

For RQ2, we used the data on the code blocks and their categories to perform statistical analysis of applied programming abstractions and concepts. Similarly to \cite{Dasgupta_2016} and \cite{Maloney_2008}, we consider the use of certain blocks to indicate that a programming abstraction or concept is being used in a certain project. In Section \ref{RQ2} we present the results related to the utilization of procedures, variables, loops, conditional statements, user interactivity and synchronization.

For RQ3, we focused on four types of code smells: duplicated code, dead code, large script and large sprite. In the duplicated code smell analysis, our first step was to define what we consider a code clone in the context of Scratch programming: a script that is composed of a set of blocks of the same type connected in the same way and is repeated within or across sprites of the same project. For the identification of clones we did not take into account the values of the parameters that may be used in the blocks, so that two blocks that only differ in the values of parameters are considered to be equal. We also investigated the case of clones with the same parameter values, and we refer to them as \textit{exact clones}. The next step in the analysis was to determine the minimum size of the scripts that are considered clones instead of incidentally similar. We examined the number of detected clones for different script sizes and present the results in Figure \ref{fig:cloneslines}. Based on this distribution, we opted to adopt the number also used by the authors in \cite{moreno_automatic_2014}, which is a minimum size of five blocks per script.

@@

In the context of Scratch, we consider the long method and the large class smells as large script and large sprite smells respectively. For these two smells we use the number of blocks as the size metric. Figure \ref{fig:longmethod} presents the number of blocks in scripts and sprites of our dataset. We used these statistics to split the dataset and retrieve the top 10\% largest scripts and sprites, as is commonly done in both source code analysis~\cite{alves_deriving_2010} and analysis of end-user programming artifacts like spreadsheets~\cite{hermans2015detecting}. Using that strategy, we set the thresholds for the calculation of the large script and large sprite: 18 blocks and 59 blocks respectively. The results we obtained using these thresholds are presented in Section \ref{RQ3}.

@@

\subsection{Program Size and Complexity}
\label{RQ1}
The dataset contains a relatively small number of projects without any code: \nemptyPrograms~ (5.77\%). Through random manual sampling we found that in some cases these projects contains only sprites and costumes, but no code, while other projects were entirely empty apart from the Scratch cat added by default. Since these projects are empty in terms of code, we excluded them from further analysis, leaving the final number of analyzed non-empty projects to \nScriptPrograms.

\begin{table*}[ht]
	\centering
	\begin{tabular}{llrrrrrr}
		&&\textbf{mean}&\textbf{min}&\textbf{Q1}&\textbf{median}&\textbf{Q3}&\textbf{max}\\
		\hline
		Size&Sprites with code per project&5.68&1&1&2&5&525\\
		&Scripts per project&17.35&1&2&5&12&3,038\\
		&Number of blocks per project&154.55&1&12&29&76&34,622\\
		&Blocks in Stage per project&4.80&0&0&0&3&2,613\\
		&Blocks in Sprites per project&115.57&0&10&26&68&34,613\\
		&Blocks in Procedures per project&34.17&0&0&0&0&20,552\\
		\hline
		Complexity&McCabe Cyclomatic Complexity (CC) per script&1.58&1&1&1&1&246\\
		&McCabe CC per procedure script&3.75&1&1&2&4&183\\
		\hline
		Procedures&Procedures per project with procedures&11.50&1&1&2&6&847\\
		&Arguments per Procedure&0.95&0&0&0&1&53\\
		&Numerical arguments per procedure with arguments&1.73&0&1&1&2&22\\
		&Text arguments per procedure with arguments&0.28&0&0&0&1&24\\
		&Boolean arguments per procedure with arguments&0.13&0&0&0&0&14\\
		&Calls per procedure&2.14&0&1&1&2&526\\
		&Scripts with calls per procedure&1.13&0&1&1&1&59\\
		\hline
		Programming&Variables per project&2.06&0&0&0&1&340\\
		concepts&Scripts utilizing variable&4.97&1&1&3&5&1,127\\
		&Lists per project&0.55&0&0&0&0&319\\
		&Conditional statements per project&10.02&0&0&0&3&5,950\\
		&Loop statements per project&7.65&0&1&2&5&2,503\\
		&User input blocks per project&4.77&0&0&1&4&1,889\\
		&Broadcast-receive statements per project&8.57&0&0&0&2&2,460\\	
		\hline
	\end{tabular}
	\caption{Summary statistics of the dataset of \nScriptPrograms~non-empty Scratch projects}
	\label{tbl-size}
\end{table*}

@@

The number of blocks metric was further analyzed to understand code organization in more depth. The majority of Scratch code---74.78\% out of \nLOC~blocks---is written within sprites. An additional 3.1\% of the total blocks are found in the stage class. More interestingly, the remaining 22.11\% are blocks within defined procedures, which are found in only 7.7\% (17,979) of the projects. The projects that contain procedures use them a lot; almost half of their total blocks (48.81\%) are within procedures.

@@

\noindent
\fbox{
	\begin{minipage}{0.46\textwidth}
		RQ1: The majority of Scratch projects are small and simple: 75\% of the projects have up to 5 sprites, 12 scripts and 76 blocks. The majority (78\%) of scripts contain no decision points. Most code is written in sprites. There exist surprisingly large and complex projects.
	\end{minipage}}

\subsection{Programming Abstractions and Concepts}
\label{RQ2}
The first method for abstraction that we investigate are procedures. In the dataset we found 206,799 procedures in 17,979 (7.7\%) projects. As summarized in Table \ref{tbl-size}, the projects that contain procedures have an average of 11.5 procedures, with 53.59\% of these projects having up to 2. Figure \ref{fig:proceduresperproject} shows the distribution of procedures in projects. Regarding procedure arguments, we found that 55.57\% have no arguments and 19.48\% have only one (shown in Figure \ref{fig:procedurearguments}). The majority of procedure arguments (80.59\%) are numeric, and the least used argument type is the boolean one---6.23\% of the total procedure arguments, found in 5.32\% of the procedures.

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/6proceduresperproject}
	\vspace{-1em}
	\caption{Number of procedures for the 17,979 projects that include at least one procedure}
	\label{fig:proceduresperproject}
\end{figure}

@@

The use of procedures in projects was further investigated through the use of procedure calls, summarized in Figure \ref{fig:procedurecalls}. Most procedures are called exactly once (62.32\% of them) or twice (14.30\%) and from exactly one script (85.92\% of them). Examining the origin of procedure calls, we observed that most of the calls (56.09\%) originate from other procedures, and 1.06\% even originate from the same procedure, making them recursive calls. These recursive procedures are found in 1,052 projects, whose identifiers are made available.\footnotemark[\ref{repo}]

@@

As shown in Table \ref{tbl-characteristics}, almost one-third of the projects use variables and a small number (4.01\%) use lists. The number of variables that is being used is also limited, with only 7.48\% of the projects having five or more variables. The distribution of variable and list utilization is shown in Figure \ref{fig:variables}. Exceptional cases exist: the analysis highlighted 842 projects with more than 100 variables and with a maximum of 340. Examining the initialization of variables through the \texttt{set <variable> to <value>} blocks, we found that for 4.83\% of all variables this was missing. While failing to initialize a variable in Scratch will not result in a runtime error as in some other programming languages, correctly setting the initial state of the program is important \cite{boe_hairball:_2013}.

@@

Regarding program control features, conditionals (\texttt{if <condition> then} and \texttt{if <condition> then else} blocks) are used by 39.81\% of the projects. Loops (blocks \texttt{repeat <tim\-es>}, \texttt{forever} and \texttt{repeat until <condition>}) are more common, used by 77.18\% of the projects. The most common of the three is the \texttt{forever} block, accounting for 51.86\% of all loops and the least common one is the \texttt{repeat until <condition>} block, accounting for 11.57\% of all loops and used in 13.59\% of the total projects.

Investigating user interactivity functionality, we found that 56.24\% of the projects in the dataset contain user input blocks---an average of 8.48 blocks per such project. Table \ref{tbl-userinput} lists the frequency of use of user input controls. We do not include the \texttt{when Green flag pressed} block here, as this is just used to start a Scratch program and hence cannot be considered input into the program. The most commonly used user input block is the \texttt{when key pressed}, found in 71,096 (30.45\% of the total) projects. The most frequently used parameter for the \texttt{key} attribute is the space key, followed by the arrows and then the letters and numbers.

@@

As detailed in Section \ref{sec:scratch}, users can define their own events, using the blocks \texttt{broadcast}, \texttt{broadcast and wait} and \texttt{when I receive}.These blocks are used in 29.57\% of the projects. \texttt{broadcast and wait} is rarely used, in only 3.87\% of the projects.

@@

We further inspected which of the identified clones were duplicated only within the same sprite: 63,682 (37.34\% of the total) clones, in 10.14\% of all projects. 2.12\% of the projects contain cloned procedure definitions, which were measured to 12,878 (7.55\% of the total) clones.

@@

Apart from whole scripts, we also examined cases where scripts differed only in the first (Hat) block. This way we examine if Scratch programmers assign the same functionality to handle different types of events. Cloned functionality blocks are found to be rare: without considering the first block, only 2,243 additional clones were found in 920 projects.

The second smell that we examine is the dead code smell. We identify four types of dead code: (1) procedures that are not invoked, (2) unmatched broadcast-receive messages, (3) code that is not invoked and (4) empty event scripts, i.e., scripts that contain an event block alone. Investigating the first type, we find that a significant number of the defined procedures (13,036 or 5.06\%) are not called in the projects. This is also shown in Figure \ref{fig:procedurecalls} and it occurs in 2,079 projects. For the second type, we examined the broadcast-receive messages and found that they are not always synchronized: 3,33\% of the \texttt{when I receive} blocks were found to wait for a message that is never being broadcasted, while 4,4\% of the \texttt{broadcast} blocks broadcast a message that is not being received. This lack of syncronization occurs in 18,669 (7.99\% of the total) projects.

The third and fourth cases are incomplete scripts. Those are either never invoked due to the lack of a starting \texttt{when <trigger>} block from the Scratch Events or Control category, or are comprised of only a \texttt{when <trigger>} block without any functionality. A total of 322,475 scripts like that were found in 56,890 (24.36\% of the total) projects. The majority of these scripts (86.6\%) are scripts missing a Hat block. Examining the size of these dead scripts, 72.34\% are composed of a single block. As shown in Figure \ref{fig:deadcode}, however, considerably large dead scripts exist; 2,358 of these scripts originating from 1,553 different projects have more than 30 blocks and up to 2,610.\footnotemark[\ref{repo}]

@@

\section{Discussion}
\label{sec:discussion}
\subsection{Implications}
We believe a large scale study of programs like the current can help language designers to tailor their language. In this section we highlight directions in which our study could support language design. There are many other implications to be considered, which is why we have made our dataset public.

@@

In a sense, the dead scripts are harmless, as they are not executed. However, they do cause `visual clutter' and might be distracting to novice programmers, as it might be hard to see which scripts are dead. In contrast to other visual educational languages, Scratch does not indicate scripts that are dead. LEGO Mindstorms, for example, does give the user feedback on dead blocks by making them gray.

Looking at the number of unconnected blocks, we hypothesize that Scratch programmers have a need for a separate workspace to store unconnected blocks temporarily. We envision that would be like the `backpack', a Scratch feature meant to move scripts across sprites. In order to help novice programmers keep their code clean, the programming interface could actively encourage users to move unconnected blocks to that workspace when they exit the environment.

\subsubsection{Exact clones between sprites}
\label{implicationsClones}
With occurrences in 11\% of the Scratch projects in our dataset, the use of exactly identical clones between sprites is relatively common. In a sense, the Scratch users are not to blame for that, as Scratch does not support procedure calls between sprites, only within them. So in many cases there is no way to share the functionality other than by making a copy. We are not aware of the underlying rationale of the Scratch team that lead to this decision, however it seems that a large part of the Scratch users would use the functionality to call procedures between sprites.

@@

\subsection{Threats to validity}
A threat to the validity of this study is the fact that we did not scrape a random sample, but the most recent 250,000 projects. It could be the case that the programming habits of Scratch users are changing over time. However, we counterbalanced that by using a large dataset which comprises around 2\% of all 14 million shared Scratch projects\footnote{\url{https://scratch.mit.edu/statistics/}}.
